---
title: "Piped Operations (the Composition of Functions)"
author: Benjamin Nutter
date: 20 November, 2015
output: ioslides_presentation
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(magrittr)
```

## Performing Complex Operations

Question:

How do we translate $g \circ f (x) = g(f(x))$ into code?

## Quick Disclaimer:

These slide use the functions `subtract`, `divide`, and `power`, which are custom aliases for `-`, `/`, and `^`, respectively.  They are defined here for convenience of instruction.

```{r}
subtract <- function(x, y){ 
  x - y 
}

divide <- function(x, y) {
  x / y
}

power <- function(x, y) {
  x ^ y
}
```

## Examples

The following code snippets all do the same calculation.  
Can you identify it?  
Which is easiest to read?

## Examples
```{r}
a <- rnorm(100, 50, 15)

solution <- a - mean(a)
solution <- sum(a)
solution <- solution^2
solution <- solution / (length(a) - 1)
solution <- sqrt(solution)

solution <- sqrt(sum((a - mean(a))^2) / (length(a) - 1))

solution <- a %>%
  subtract(mean(a)) %>%
  power(2) %>%
  sum() %>%
  divide(length(a) - 1) %>%
  sqrt()
```

## Answer

$$\sqrt{\frac{\sum\limits_{i=1}^{n}(x_i - \bar{x})^2}{n-1}}$$

Which one looked easiest to you?

## Pipes Vs. Step-by-Step
```{r}
a <- rnorm(100, 50, 15)

solution <- a - mean(a)
solution <- sum(a)
solution <- solution^2
solution <- solution / (length(a) - 1)
solution <- sqrt(solution)

solution <- a %>%
  subtract(mean(a)) %>%
  power(2) %>%
  sum() %>%
  divide(length(a) - 1) %>%
  sqrt()
```

## Pipes vs. Nested Functions
```{r}
solution <- sqrt(sum((a - mean(a))^2) / (length(a) - 1))

solution <- a %>%
  subtract(mean(a)) %>%
  power(2) %>%
  sum() %>%
  divide(length(a) - 1) %>%
  sqrt()
```

## The Crash Course in Piping

The fully specified form of a pipe is

```{r}
solution <- a %>%
  subtract(., mean(a)) %>%
  power(., 2) %>%
  sum(.) %>%
  divide(., length(a) - 1) %>%
  sqrt(.)
```

Notice all of those dots!

## The Crash Course in Piping

The `.` is a place holder in the pipe representing "The result from the previous command"

```{r}
solution <- a %>%
  subtract(., mean(a)) %>%
  power(., 2) %>%
  sum(.) %>%
  divide(., length(a) - 1) %>%
  sqrt(.)
```

## The Crash Course in Piping

When the result of the previous command will go in the first argument of the subsequent command, the `.` is optional.

```{r}
solution <- a %>%
  subtract(mean(a)) %>%
  power(2) %>%
  sum() %>%
  divide(length(a) - 1) %>%
  sqrt()
```

## The Crash Course in Piping

When the result of the previous command will go into any other argument of the subsequent command, the dot is mandatory.

Compare

```{r}
a <- 50

a %>%
  subtract(25)

a %>%
  subtract(25, .)
```

